### 2010-02-23 ChrisnaSections.tex

    %% RESULTS:
    \subsection{Results}
    Below, in tables, lie our estimates for all of the Minnesota DNR's data from
    2003 and 2004.  Below each table is detailed commentary about the ability of
    the standard hierarchical model to convergence to the corresponding data. \\

    % Mississippi River!
    \noindent{\begin{tabular}{|l|l|l|l|}
        \hline
        \multicolumn{4}{|l|}{\textbf{Mississippi River, 2003}} \\
        \hline
            & 402m & 804m & 1608m \\
        \hline
        \multirow{2}{*}{E[Danny]}
            & 0.01954 & 0.04808 & 0.06885 \\
            & (0.006515, 0.03663) & (0.01874, 0.08187) & (0.01778, 0.1383) \\
        \hline
        \multirow{2}{*}{E[John]}
            & 0.05216 & 0.07771 & 0.13650 \\
            & (0.027710, 0.07745) & (0.03881, 0.12100) & (0.03449, 0.2330) \\
        \hline
        \multirow{2}{*}{E[Tom]}
            & 0.02668 & 0.06140 & 0.10230 \\
            & (0.011990, 0.04622) & (0.02789, 0.10390) & (0.04287, 0.1796) \\
        \hline
        \multirow{2}{*}{p[Danny]}
            & 0.65240 & 0.67340 & 0.74960 \\
            & (0.548300, 0.75990) & (0.56760, 0.78290) & (0.60510, 0.8910) \\
        \hline
        \multirow{2}{*}{p[John]}
            & 0.67770 & 0.77580 & 0.82540 \\
            & (0.583500, 0.77280) & (0.66060, 0.88690) & (0.71390, 0.9326) \\
        \hline
        \multirow{2}{*}{p[Tom]}
            & 0.53670 & 0.54220 & 0.65330 \\
            & (0.501200, 0.61300) & (0.50160, 0.62850) & (0.51530, 0.8104) \\
        \hline
        \multirow{2}{*}{\(\psi\)}
            & 0.59380 & 0.69370 & 0.75820 \\
            & (0.439000, 0.75990) & (0.51800, 0.87380) & (0.54110, 0.9300) \\
        \hline
        \multirow{2}{*}{\(\theta\)}
            & 0.15280 & 0.20730 & 0.27690 \\
            & (0.111900, 0.19780) & (0.14580, 0.27390) & (0.19030, 0.3716) \\
        \hline
    \end{tabular}}
    \\

    This data converged well at the 402m and 804m plot sizes.  At the 1608m plot
    size, \(E\), \(\psi\), and \(\theta\) did not converge well. \\

    \noindent{\begin{tabular}{|l|l|}
        \hline
        \multicolumn{2}{|l|}{\textbf{Mississippi River, 2004}} \\
        \hline
            & 402m \\
        \hline
        \multirow{2}{*}{p[Danny]}
            & 0.3129 \\
            & (0.2174, 0.8107) \\
        \hline
        \multirow{2}{*}{\(\psi\)}
            & 0.7613 \\
            & (0.6609, 0.8833) \\
        \hline
        \multirow{2}{*}{\(\theta\)}
            & 0.8511 \\
            & (0.2426, 0.9973) \\
        \hline
    \end{tabular}}
    \\

    This data did not converge.  We do not think that these estimates are
    accurate, but they are included for the sake of completeness. \\

    % Whitewater River!
    \noindent{\begin{tabular}{|l|l|l|}
        \hline
        \multicolumn{3}{|l|}{\textbf{Whitewater River, 2003}} \\
        \hline
            & 402m & 804m \\
        \hline
        \multirow{2}{*}{E[Danny]}
            & 0.05608 & 0.0555 \\
            & (0.02474, 0.09554) & (0.002859, 0.1390) \\
        \hline
        \multirow{2}{*}{E[John]}
            & 0.16570 & 0.1931 \\
            & (0.10130, 0.23510) & (0.065910, 0.3384) \\
        \hline
        \multirow{2}{*}{E[Tom]}
            & 0.15350 & 0.2241 \\
            & (0.09758, 0.21660) & (0.110000, 0.3623) \\
        \hline
        \multirow{2}{*}{p[Danny]}
            & 0.66990 & 0.6985 \\
            & (0.56080, 0.79040) & (0.573100, 0.8366) \\
        \hline
        \multirow{2}{*}{p[John]}
            & 0.92040 & 0.9877 \\
            & (0.84270, 0.97590) & (0.945100, 1.0000) \\
        \hline
        \multirow{2}{*}{p[Tom]}
            & 0.75910 & 0.8575 \\
            & (0.65060, 0.85800) & (0.745400, 0.9600) \\
        \hline
        \multirow{2}{*}{\(\psi\)}
            & 0.87590 & 0.9126 \\
            & (0.72910, 0.98880) & (0.746900, 0.9955) \\
        \hline
        \multirow{2}{*}{\(\theta\)}
            & 0.24100 & 0.3779 \\
            & (0.17890, 0.30900) & (0.286500, 0.4721) \\
        \hline
    \end{tabular}}
    \\

    While the 402m data converged well, \(E\), \(\psi\), and \(\theta\) did not
    converge well for the 804m data. \\

    \noindent{\begin{tabular}{|l|l|l|l|}
        \hline
        \multicolumn{2}{|l|}{\textbf{Whitewater River, 2004}} \\
        \hline
            & 402m \\
        \hline
        \multirow{2}{*}{p[Danny]}
            & 0.8237 \\
            & (0.5323, 0.9804) \\
        \hline
        \multirow{2}{*}{\(\psi\)}
            & 0.6066 \\
            & (0.4246, 0.8382) \\
        \hline
        \multirow{2}{*}{\(\theta\)}
            & 0.2068 \\
            & (0.1211, 0.3550) \\
        \hline
    \end{tabular}}
    \\

    This data converged well.  However, we feel that the absence of \(E\) from
    the model makes the estimates much less dependable.  The estimates are also
    much less precise than those from 2003 -- the confidence intervals of
    \(\psi\) and of \(p[Danny]\) are twice as large as those from 2003. \\

    % Zumbro River!
    \noindent{\begin{tabular}{|l|l|l|}
        \hline
        \multicolumn{3}{|l|}{\textbf{Zumbro River, 2003}} \\
        \hline
            & 402m & 804m \\
        \hline
        \multirow{2}{*}{E[Danny]}
            & 0.02922 & 0.007066 \\
            & (0.006786, 0.05994) & (0.0000, 0.03403) \\
        \hline
        \multirow{2}{*}{E[John]}
            & 0.03388 & 0.008312 \\
            & (0.008584, 0.06880) & (0.0000, 0.04188) \\
        \hline
        \multirow{2}{*}{E[Tom]}
            & 0.02228 & 0.012960 \\
            & (0.002430, 0.05385) & (0.0000, 0.05642) \\
        \hline
        \multirow{2}{*}{p[Danny]}
            & 0.81890 & 0.837100 \\
            & (0.749500, 0.88540) & (0.7615, 0.90700) \\
        \hline
        \multirow{2}{*}{p[John]}
            & 0.81770 & 0.852500 \\
            & (0.740100, 0.88670) & (0.7597, 0.92230) \\
        \hline
        \multirow{2}{*}{p[Tom]}
            & 0.73120 & 0.805400 \\
            & (0.625400, 0.82440) & (0.6885, 0.89780) \\
        \hline
        \multirow{2}{*}{\(\psi\)}
            & 0.66550 & 0.651600 \\
            & (0.549100, 0.78150) & (0.5098, 0.78210) \\
        \hline
        \multirow{2}{*}{\(\theta\)}
            & 0.29950 & 0.395800 \\
            & (0.236400, 0.36550) & (0.3100, 0.48520) \\
        \hline
    \end{tabular}}
    \\

    Again, the 402m data converged well.  With respect to the 804m data, while
    \(\psi\) and \(\theta\) converged well, none of the other parameters did. \\

    \noindent{\begin{tabular}{|l|l|l|l|}
        \hline
        \multicolumn{2}{|l|}{\textbf{Zumbro River, 2004}} \\
        \hline
            & 402m \\
        \hline
        \multirow{2}{*}{p[Danny]}
            & 0.9166 \\
            & (0.7739, 0.9918) \\
        \hline
        \multirow{2}{*}{\(\psi\)}
            & 0.4773 \\
            & (0.3586, 0.6114) \\
        \hline
        \multirow{2}{*}{\(\theta\)}
            & 0.4232 \\
            & (0.3073, 0.5520) \\
        \hline
    \end{tabular}}
    \\

>>>>

    %% RESULTS:
    \subsection{Results}
    Below, in tables, lie our estimates for all of the Minnesota DNR's data from
    2003 and 2004.  Below each table is detailed commentary about the ability of
    the standard hierarchical model to convergence to the corresponding data.

    With respect to the model's ability to converge on the actual data, the most
    important factors are plot size, the number of flights recorded, and the
    number of estimated parameters.  In general, both plot size and the number
    of flights recorded must be sufficiently small and large, respectively, for
    the standard model with \(E\) to converge.

    2003's data worked relatively well with our model.  The presence of numerous
    observations that occur in the second and third days since the most recent
    snow event allows one of the more crucial assumptions of our hierarchical
    model -- that otter tracks laid on each day after a snow event are visible
    throughout the entire observation period and therefore, that it is more
    likely for an observer to detect a track on later days -- to be thoroughly
    tested.

    2004's data, which suffered from the lack of multiple observers and
    observations later than one day after each snow event, does not provide
    enough data with respect to time (measured by the number of days since the
    last snow event) for the parameters \(p\) and \(E\) to converge, regardless
    of the number of Markov Chain Monte Carlo (MCMC) iterations performed.

    % Mississippi River!
    \begin{center}
    \begin{tabular}{|l|l|l|l|}
        \hline
        \multicolumn{4}{|l|}{\textbf{Mississippi River, 2003}} \\
        \hline
            & 402m & 804m & 1608m \\
        \hline
        \multirow{2}{*}{E[Danny]}
            & 0.01954 & 0.04808 & 0.06885 \\
            & (0.006515, 0.03663) & (0.01874, 0.08187) & (0.01778, 0.1383) \\
        \hline
        \multirow{2}{*}{E[John]}
            & 0.05216 & 0.07771 & 0.13650 \\
            & (0.027710, 0.07745) & (0.03881, 0.12100) & (0.03449, 0.2330) \\
        \hline
        \multirow{2}{*}{E[Tom]}
            & 0.02668 & 0.06140 & 0.10230 \\
            & (0.011990, 0.04622) & (0.02789, 0.10390) & (0.04287, 0.1796) \\
        \hline
        \multirow{2}{*}{p[Danny]}
            & 0.65240 & 0.67340 & 0.74960 \\
            & (0.548300, 0.75990) & (0.56760, 0.78290) & (0.60510, 0.8910) \\
        \hline
        \multirow{2}{*}{p[John]}
            & 0.67770 & 0.77580 & 0.82540 \\
            & (0.583500, 0.77280) & (0.66060, 0.88690) & (0.71390, 0.9326) \\
        \hline
        \multirow{2}{*}{p[Tom]}
            & 0.53670 & 0.54220 & 0.65330 \\
            & (0.501200, 0.61300) & (0.50160, 0.62850) & (0.51530, 0.8104) \\
        \hline
        \multirow{2}{*}{\(\psi\)}
            & 0.59380 & 0.69370 & 0.75820 \\
            & (0.439000, 0.75990) & (0.51800, 0.87380) & (0.54110, 0.9300) \\
        \hline
        \multirow{2}{*}{\(\theta\)}
            & 0.15280 & 0.20730 & 0.27690 \\
            & (0.111900, 0.19780) & (0.14580, 0.27390) & (0.19030, 0.3716) \\
        \hline
    \end{tabular}
    \end{center}

    This data converged well at the 402m and 804m plot sizes.  At the 1608m plot
    size, \(E\), \(\psi\), and \(\theta\) did not converge well.

    \begin{center}
    \begin{tabular}{|l|l|}
        \hline
        \multicolumn{2}{|l|}{\textbf{Mississippi River, 2004}} \\
        \hline
            & 402m \\
        \hline
        \multirow{2}{*}{p[Danny]}
            & 0.3129 \\
            & (0.2174, 0.8107) \\
        \hline
        \multirow{2}{*}{\(\psi\)}
            & 0.7613 \\
            & (0.6609, 0.8833) \\
        \hline
        \multirow{2}{*}{\(\theta\)}
            & 0.8511 \\
            & (0.2426, 0.9973) \\
        \hline
    \end{tabular}
    \end{center}

    This data did not converge.  We do not think that these estimates are
    accurate, but they are included for the sake of completeness.

    % Whitewater River!
    \begin{center}
    \begin{tabular}{|l|l|l|}
        \hline
        \multicolumn{3}{|l|}{\textbf{Whitewater River, 2003}} \\
        \hline
            & 402m & 804m \\
        \hline
        \multirow{2}{*}{E[Danny]}
            & 0.05608 & 0.0555 \\
            & (0.02474, 0.09554) & (0.002859, 0.1390) \\
        \hline
        \multirow{2}{*}{E[John]}
            & 0.16570 & 0.1931 \\
            & (0.10130, 0.23510) & (0.065910, 0.3384) \\
        \hline
        \multirow{2}{*}{E[Tom]}
            & 0.15350 & 0.2241 \\
            & (0.09758, 0.21660) & (0.110000, 0.3623) \\
        \hline
        \multirow{2}{*}{p[Danny]}
            & 0.66990 & 0.6985 \\
            & (0.56080, 0.79040) & (0.573100, 0.8366) \\
        \hline
        \multirow{2}{*}{p[John]}
            & 0.92040 & 0.9877 \\
            & (0.84270, 0.97590) & (0.945100, 1.0000) \\
        \hline
        \multirow{2}{*}{p[Tom]}
            & 0.75910 & 0.8575 \\
            & (0.65060, 0.85800) & (0.745400, 0.9600) \\
        \hline
        \multirow{2}{*}{\(\psi\)}
            & 0.87590 & 0.9126 \\
            & (0.72910, 0.98880) & (0.746900, 0.9955) \\
        \hline
        \multirow{2}{*}{\(\theta\)}
            & 0.24100 & 0.3779 \\
            & (0.17890, 0.30900) & (0.286500, 0.4721) \\
        \hline
    \end{tabular}
    \end{center}

    While the 402m data converged well, \(E\), \(\psi\), and \(\theta\) did not
    converge well for the 804m data.

    \begin{center}
    \begin{tabular}{|l|l|l|l|}
        \hline
        \multicolumn{2}{|l|}{\textbf{Whitewater River, 2004}} \\
        \hline
            & 402m \\
        \hline
        \multirow{2}{*}{p[Danny]}
            & 0.8237 \\
            & (0.5323, 0.9804) \\
        \hline
        \multirow{2}{*}{\(\psi\)}
            & 0.6066 \\
            & (0.4246, 0.8382) \\
        \hline
        \multirow{2}{*}{\(\theta\)}
            & 0.2068 \\
            & (0.1211, 0.3550) \\
        \hline
    \end{tabular}
    \end{center}

    This data converged well.  However, we feel that the absence of \(E\) from
    the model makes the estimates much less dependable.  The estimates are also
    much less precise than those from 2003 -- the confidence intervals of
    \(\psi\) and of \(p[Danny]\) are twice as large as those from 2003.

    % Zumbro River!
    \begin{center}
    \begin{tabular}{|l|l|l|}
        \hline
        \multicolumn{3}{|l|}{\textbf{Zumbro River, 2003}} \\
        \hline
            & 402m & 804m \\
        \hline
        \multirow{2}{*}{E[Danny]}
            & 0.02922 & 0.007066 \\
            & (0.006786, 0.05994) & (0.0000, 0.03403) \\
        \hline
        \multirow{2}{*}{E[John]}
            & 0.03388 & 0.008312 \\
            & (0.008584, 0.06880) & (0.0000, 0.04188) \\
        \hline
        \multirow{2}{*}{E[Tom]}
            & 0.02228 & 0.012960 \\
            & (0.002430, 0.05385) & (0.0000, 0.05642) \\
        \hline
        \multirow{2}{*}{p[Danny]}
            & 0.81890 & 0.837100 \\
            & (0.749500, 0.88540) & (0.7615, 0.90700) \\
        \hline
        \multirow{2}{*}{p[John]}
            & 0.81770 & 0.852500 \\
            & (0.740100, 0.88670) & (0.7597, 0.92230) \\
        \hline
        \multirow{2}{*}{p[Tom]}
            & 0.73120 & 0.805400 \\
            & (0.625400, 0.82440) & (0.6885, 0.89780) \\
        \hline
        \multirow{2}{*}{\(\psi\)}
            & 0.66550 & 0.651600 \\
            & (0.549100, 0.78150) & (0.5098, 0.78210) \\
        \hline
        \multirow{2}{*}{\(\theta\)}
            & 0.29950 & 0.395800 \\
            & (0.236400, 0.36550) & (0.3100, 0.48520) \\
        \hline
    \end{tabular}
    \end{center}

    Again, the 402m data converged well.  With respect to the 804m data, while
    \(\psi\) and \(\theta\) converged well, none of the other parameters did.

    \begin{center}
    \begin{tabular}{|l|l|l|l|}
        \hline
        \multicolumn{2}{|l|}{\textbf{Zumbro River, 2004}} \\
        \hline
            & 402m \\
        \hline
        \multirow{2}{*}{p[Danny]}
            & 0.9166 \\
            & (0.7739, 0.9918) \\
        \hline
        \multirow{2}{*}{\(\psi\)}
            & 0.4773 \\
            & (0.3586, 0.6114) \\
        \hline
        \multirow{2}{*}{\(\theta\)}
            & 0.4232 \\
            & (0.3073, 0.5520) \\
        \hline
    \end{tabular}
    \end{center}

### 2010-02-23 kivaSections.tex

In 2003, three different observers flew over the rivers. Assuming that no tracks are laid down between individuals’ flights on the same day, all three people should have seen the same tracks. While there is some error in classifying which site the observer saw the actual track in, each observer should at least be seeing the same, or nearly the same, number of sites with tracks. However, we did not find that to be the case. If we look only at days that all three people observed, there appear to be major differences among the proportion of sites in which the scientists detected a track (Figure x). The difference in estimated na\"ive occupancy among the observers indicates that the scientists did not always correctly identify the tracks.

@@

\subsection{False positive rates}
The lack of consistency among the flights is strong evidence that the data contain false positives where scientists think they see a track when it does not exist. Not accounting for such a process can lead to major biases; as the number of flights increases, the occupancy rate estimate will approach one (Royle and Link 2006). We used the observer as a covariate (by means of indicator variables) for both the probability of correctly identifying a track that exists ($p$) and the probability of identifying a track that does not exist ($E$). Including false positives introduces a duality whereby $(p = x, E = y, \psi = z)$ is as good a solution as $(p = y, E = x, \psi = 1-z)$. One possible way to address this problem is to assume $p\gg0.5$ and $E\ll0.5$.

@@

>>>>

In 2003, three different observers flew over the rivers. Assuming that no tracks 
are laid down between individuals’ flights on the same day, all three people 
should have seen the same tracks. While there is some error in classifying which 
site the observer saw the actual track in, each observer should at least be 
seeing the same, or nearly the same, number of sites with tracks. However, we 
did not find that to be the case. If we look only at days that all three people 
observed, there appear to be major differences among the proportion of sites in 
which the scientists detected a track (Figure x). The difference in estimated na
\"ive occupancy among the observers indicates that the scientists did not always 
correctly identify the tracks.

@@

\subsection{False positive rates}
The lack of consistency among the flights is strong evidence that the data 
contain false positives where scientists think they see a track when it does not 
exist. Not accounting for such a process can lead to major biases; as the number 
of flights increases, the occupancy rate estimate will approach one (Royle and 
Link 2006). We introduced a false detection process and used the observer as a 
covariate (by means of indicator variables) for both the probability of 
correctly identifying a track that exists ($p$) and the probability of 
identifying a track that does not exist ($E$). Including false positives 
introduces a duality whereby $\begin{array}{ccc}(p = x, & E = y, & \psi = z)\end
{array}$ is as good a solution as $\begin{array}{ccc}(p = y, & E = x, & \psi = 
1-z)\end{array}$. To address this problem we assumed $p\gg0.5$ and $E\ll0.5$, an 
approach recommended by Royle and Link (2006).

@@

\section{Spatial correlation of the data}
We found that there is significant correlation between the probability of 
observing a track in neighboring sites, but this violates of a central 
assumption of occupancy theory. This issue needed to be addressed. One approach 
was to make the prior distribution of $\theta$, the probability that an otter 
lays down a track, spatially correlated using a conditional autoregressive
(CAR) model.
